Use of significance test logic by scientists in a novel reasoning task

### 2025-03-07 inst/manuscript/manuscript.Rmd
small changes to prepare for resubmission

abstract: |
  Although statistical significance testing is one of the most widely-used techniques across science, previous research has suggested that scientists have a poor understanding of how it works. If scientists misunderstand one of their primary inferential tools the implications are dramatic: potentially unchecked, unjustified conclusions and wasted resources. Scientists' apparent difficulties with significance testing have led to calls for its abandonment or increased reliance on alternative tools, which would represent a substantial, untested, shift in scientific practice. However, if scientists' understanding of significance testing is truly as poor as thought, one could argue such drastic action is required. We present evidence using a novel reasoning task that scientists may understand the logic of significance testing better than previously thought. Scientists may not be as statistically-challenged as often believed; reforms should take this into account.

@@

The interpretation of studies of researchers' understanding of SST is limited, however, by their methodology. A typical study presents a vignette describing research results. Statistical results are offered to the participants (e.g., a $t$ statistic and $p$ value), who are then asked to explicitly give or endorse various interpretations. These responses are taken to represent participants' understanding, or misunderstanding, of SST. However, there are reasons to be cautious of drawing strong conclusions from these studies, including the abstract nature of such vignettes, the lack of investment researchers have in the fictional research, and their disconnection from research activity (e.g., experimentation and replication). It is unclear how well vignette studies (including ones by the present authors: [@Hoekstra:etal:2014;@Hoekstra:etal:2012]) tap understanding of the core logic of SST rather than, say, familiarity with the technical terminology used to present statistical results. Conceptual understanding and fluency with common representations are both important, but are distinct.

@@

If participants were using the shuffle reports to interpret the data, as would be predicted if they were using SST logic, the transformation should affect their interpretation of the visual evidence: a visually-extreme point should be more discounted against the sampling distribution that is wider. When we break down responses by the *visual* extremeness of the evidence, responses in two conditions should appear different; when we break down responses by  *statistical* extremeness (i.e., the $p$ value) responses in the two conditions should appear very similar, because the visual manipulation is irrelevant given the $p$ value.

@@

We wish to emphasize what we cannot, and do not, argue. First, we cannot argue that simply because scientists can successfully use SST logic, that they *do* in real situations, or that specific instantiations of SST, such as $p$ values, are used well. We specifically set out to abstract the logic away from the typical situations in which scientists use the logic. This has the benefit of being helping to identify where problems might be, but the downside that generalizing the results will require further work. We also cannot address other potential arguments against SST, such as philosophical ones.

Finally, we hope to provide a fresh method and perspective on a long-standing debate in statistical cognition. Simulation-based approaches to teaching statistics have long been touted [@Cumming:etal:1995;@Rossman:Chance:2014]. Simulation-based approaches to *studying* scientists' statistical reasoning may also profitable, particularly in studying reasoning that is difficult for participants to articulate formally. If we are to reform statistical education and practice in the sciences, we should base that reform on diverse lines of evidence about scientists' reasoning. Understanding and harnessing scientists' already-existing competence in statistical reasoning is essential to developing effective methodological reforms.

>>>>

abstract: |
  Although statistical significance testing is one of the most widely-used techniques across science, previous research has suggested that scientists have a poor understanding of how it works. If scientists misunderstand one of their primary inferential tools the implications are dramatic: potentially unchecked, unjustified conclusions and wasted resources. Scientists' apparent difficulties with significance testing have led to calls for its abandonment or increased reliance on alternative tools, which would represent a substantial, untested, shift in scientific practice. However, if scientists' understanding of significance testing is truly as poor as thought, one could argue such drastic action is required. We present evidence using a novel reasoning task that scientists successfully use the logic of significance testing, which has implications for statistical reform.

@@

The interpretation of studies of researchers' understanding of SST is limited, however, by their methodology. A typical study presents a vignette describing research results. Statistical results are offered to the participants (e.g., a $t$ statistic and $p$ value), who are then asked to explicitly give or endorse various interpretations. These responses are taken to represent participants' understanding, or misunderstanding, of SST. However, there are reasons to be cautious of drawing strong conclusions from these studies, including the abstract nature of such vignettes, the lack of investment researchers have in the fictional research, and their disconnection from research activity (e.g., experimentation and replication). It is unclear how well vignette studies (including ones by the present authors: [@Hoekstra:etal:2014;@Hoekstra:etal:2012]) tap understanding of the core logic of SST rather than, say, the human factors of a particular practical implementation of SST (e.g., $p$ values). Conceptual understanding and fluency with common implementations are both important, but are distinct.

@@

If participants were using the shuffle reports to interpret the data, as would be predicted if they were using SST logic, the transformation should affect their interpretation of the visual evidence: a visually-extreme point should be more discounted against the sampling distribution that is wider. Visual extremeness is misleading by itself; extremeness only matters relative to the null sampling distribution. When we break down responses by the visual extremeness of the evidence, responses in two conditions should appear different; when we break down responses by  statistical extremeness (i.e., the $p$ value) responses in the two conditions should appear very similar, because the visual manipulation is irrelevant given the $p$ value.

@@

We wish to emphasize what we cannot, and do not, argue. First, we cannot argue that simply because scientists can successfully use SST logic, that they *do* in real situations, or that specific instantiations of SST, such as $p$ values, are used well. We specifically set out to abstract the logic away from the typical situations in which scientists use the logic. This has the benefit of helping to identify where problems might be, but has the downside that generalizing the results will require further work. We also cannot address other potential arguments against SST, such as philosophical ones.

Finally, we hope to provide a fresh method and perspective on a long-standing debate in statistical cognition. Simulation-based approaches to teaching statistics have long been touted [@Cumming:etal:1995;@Rossman:Chance:2014]. Simulation-based approaches to *studying* scientists' statistical reasoning may also be profitable, particularly in studying reasoning that is difficult for participants to formally articulate. If we are to reform statistical education and practice in the sciences, we should base that reform on diverse lines of evidence about scientists' reasoning. Understanding and harnessing scientists' already-existing competence in statistical reasoning is essential to developing effective methodological reforms.
